---
title: "Keystone exercise: Estuary Fish Survey - Data Rescue Mission"
author: "Kira Roberts"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## Introduction

This project was inherited from a former PhD student who was studying the relationship between predatory fish assemblages and the water quality gradient along the Ross River (Townsville, QLD).

Unfortunately for us, their data management was completed manually and the files are a disaster. Our primary task is to clean the data, and produce a master data set, from which we can produce a publication-ready visualization of the estuary's ecology.

### Housekeeping and setup

```{r setup, message=FALSE, warning=FALSE}
# Clearing the environment
rm(list = ls())

# Loading the packages needed 
library(dplyr)
library(tidyverse)
library(readxl)
library(ggplot2)
```

### The data

There were 4 datasets that were given to us. One is an excel file with the data spread over several sheets, with each needing to be loaded in separately.

```{r datadump}
# Loading in the data
catch1 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 4)

catch_data <- bind_rows(catch1, catch2, catch3, catch4)

metadata <- read.csv("../data/estuary_metadata.csv")

raw_sonde_data <- read.csv("../data/estuary_sonde_data.csv") 

species_dictionary <- read.csv("../data/species_dictionary.csv") 


# Check that all files are loaded and what we expect
head(metadata)
head(catch_data)
head(species_dictionary)
head(raw_sonde_data)

```

## The Deep Clean
Looking at the data, we can see several inconsistencies in the naming of variables and organisation of dates/times, looking deeper shows that there is also issues with the data itself where entries are missing or contain legacy errors. These need to be adjusted to ensure the datasets can be joined and accurately displayed.

### Metadata

```{r cleanmeta}
# Using glimpse() we can see the inconsistencies in site_name
glimpse(metadata) 

# Standardizing site names
clean_metadata <- metadata %>% 
  mutate(site_name = str_to_sentence(site_name), 
         site_name = str_trim(site_name), 
         site_name = str_replace_all(site_name, pattern = " ", replacement = "_")) %>% 
  rename(site = site_name)  # Rename to merge with catch data
print(clean_metadata)
```

### The Catch Data

```{r cleancatch}
# Using glimpse() we can see the inconsistencies in the species names and that the date is recognized as a date-time object. 
glimpse(catch_data)
# Because glimpse shows only one variable for site, we can use unique() to shows us all of them - and how they are inconsistent with the metadata 
unique(catch_data$site)

cleancatch_data <- catch_data %>%  
  mutate(site = str_to_sentence(site), # Standardizing site names
         site = str_trim(site), 
         site = str_replace_all(site, pattern = " ", replacement = "_")) %>%  
  mutate(species = str_to_sentence(species), # Standardizing species
         species = str_trim(species), 
         species = str_replace_all(species, pattern = " ", replacement = "_")) %>% 
  rename(common_name = species) %>% # adjusting column names to merge with species dictionary
  mutate(date = ymd(date)) %>% # Standardizing dates
  tidyr::complete(site, date, common_name, fill = list(count = 0)) #ensures that if no data was recorded, a 0 is in the place where the data should have been recorded


print(cleancatch_data)

```

### Species Information

```{r cleandictionary}
glimpse(species_dictionary)

# Standardizing species 
species_dictionary <- species_dictionary %>% 
  mutate(common_name = str_to_sentence(common_name), 
         common_name = str_trim(common_name), 
         common_name = str_replace_all(common_name, pattern = " ", replacement = "_"))

print(species_dictionary)
```

### The Environmental Data

```{r cleansonde}
# Using glimpse() we can see that the timestamp is listed as a character (<chr>) rather than a datetime object (<dttm>) and unique() shows that while the site data is more consistent than previously it still needs to be standardized
glimpse(raw_sonde_data)
unique(raw_sonde_data$site)

sensor_data <- raw_sonde_data %>%  
  mutate(site = str_to_sentence(site), # Standardizing site names
         site = str_trim(site), 
         site = str_replace_all(site, pattern = " ", replacement = "_")) %>% 
  mutate(timestamp = dmy_hm(timestamp)) %>% # Standardizing dates
  mutate(turbidity = na_if(turbidity, -999)) %>% # Clearing out bad data
  mutate(salinity = ifelse(salinity < 0, NA, salinity)) 

head(sensor_data)

# The sensor loggers collected data every 15min, realistically we don't need data at this resolution - as such we need to work out an average for each day. 
daily_sensordata <- sensor_data %>% 
  group_by(site, date = as_date(timestamp)) %>% 
  summarise(daymean_temp = mean(temperature, na.rm = TRUE),
            daymean_salinity = mean(salinity, na.rm = TRUE),
            daymean_turbidity = mean(turbidity, na.rm = TRUE),
            .groups = "drop")

head(daily_sensordata)
```

## Joining the Clean Datasets
While it is possible to join all the data in one lot of code, stepping through the process allows us to ensure that the datasets are joining how we want them to.  
```{r merge}
# Joining the Fish data
merge_catchspecies <- cleancatch_data %>% 
  left_join(species_dictionary, by = join_by(common_name))

print(merge_catchspecies)


# Joining environmental data
merge_metasensor <- daily_sensordata %>% 
  left_join(clean_metadata, by = join_by(site))

print(merge_metasensor)


# Creating one large dataset
merged_data <- merge_metasensor %>% 
  left_join(merge_catchspecies, by = join_by(date, site))

view(merged_data)
```

## Statistical Summary

A statistical summary table displaying the mean and standard deviation of fish counts and salinity for each species per zone.

```{r summary}
summary_table <- merged_data %>% 
  group_by(scientific_name, zone) %>% 
  summarise(n = n(),
            mean_count = mean(count, na.rm = TRUE),
            sd_count = sd(count, na.rm = TRUE),
            se_count = sd(count, na.rm = TRUE) / sqrt((n)),
            mean_salinity = mean(daymean_salinity, na.rm = TRUE),
            sd_salinity = sd(daymean_salinity, na.rm = TRUE),
            se_salinity = sd(daymean_salinity, na.rm = TRUE)/ sqrt ((n)),
            .groups = "drop")
print(summary_table)

```

## Plotting the Data

3.  A multi-panel plot (faceted by species) showing fish abundance along the salinity gradient, formatted to professional publication standards.

```{r, fig.cap="Fish Abundance Along the Salinity Gradient", include=FALSE}
summary_table %>% 
  ggplot() +
  geom_ribbon(aes(x = mean_salinity, ymin = mean_count - se_count, ymax = mean_count + se_count),  fill = "lightgrey") +
  geom_line(aes(x = mean_salinity, y = mean_count)) + 
  facet_wrap(~scientific_name, scales = "fixed") +
  theme_bw() + 
  theme(strip.text = element_text(face = "italic", size = 10, colour = "grey20")) +
  labs(y = "Daily Mean Abundance (+- SE)",
        x = "Salinity Gradient")
```

spatially organised
```{r, fig.cap="Fish abundance of four species along the spatial salinity gradient of an estuary"}
summary_table %>% 
  ggplot() +
  geom_point(data = merged_data, aes(x = reorder(zone, daymean_salinity), y = count), position = position_jitter(0.15), alpha= 0.1, inherit.aes = FALSE) + 
  geom_errorbar(aes(x = zone, ymin = mean_count - se_count, ymax = mean_count + se_count), width = 0.3) +
  geom_point(aes(x = reorder(zone, mean_salinity), y = mean_count, colour = zone), size = 3) +
  facet_wrap(~scientific_name, scales = "fixed") +
  theme_bw() + 
  theme(legend.position = "none") +
  theme(strip.text = element_text(face = "italic", size = 10, colour = "grey20")) +
  labs(y = "Fish Abundance",
        x = "Spacial Salinity Gradient")
```
