---
title: "Keystone exercise: Estuary Fish Survey - Data Rescue Mission"
author: "KiraRoberts"
format: html
editor: visual
---
#introduction

#housekeeping and setup
```{r, message=FALSE, warning=FALSE}
# loading packages

library(dplyr)
library(tidyverse)
library(readxl)
library(ggplot2)
```

```{r}
# Clearing the environment
rm(list = ls())

# loading data
catch1 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 4)

catch_data <- bind_rows(catch1, catch2, catch3, catch4)

head(catch_data)

metadata <- read.csv("../data/estuary_metadata.csv")
head(metadata)

raw_sonde_data <- read.csv("../data/estuary_sonde_data.csv") 
head(raw_sonde_data)

species_dictionary <- read.csv("../data/species_dictionary.csv") 
head(species_dictionary)
```
#Keystone exercise: Estuary Fish Survey - Data Rescue Mission
**The Premise:** You have inherited a project from a former PhD student who was studying the relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient. Unfortunately, their data management was entirely manual, and the files are a disaster.

They left you with four separate datasets:
1. estuary_catch_log.xlsx: Handwritten field counts of fish species (using common names), spread across multiple spreadsheet tabs.
2. estuary_metadata.csv: The spatial coordinates and estuary zones for each site.
3. estuary_sonde_data.csv: High-frequency water quality sensor readings.
4. species_dictionary.csv: A taxonomic lookup table linking local common names to their accepted scientific names.

Your mission is to write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.

**Deliverables (to include in ePortfolio report/page)**

At the end of this exercise, your final script/document must successfully output:

1. A clean master dataset containing no legacy errors (-999.0), standardised text, a complete zero-catch framework, and scientific names instead of common names.

```{r cleanmeta}
# Standardizing site names
clean_metadata <- metadata %>% 
  mutate(site_name = str_to_sentence(site_name), 
         site_name = str_trim(site_name), 
         site_name = str_replace_all(site_name, pattern = " ", replacement = "_")) %>% 
  rename(site = site_name)  # Rename to merge with catch data
print(clean_metadata)
```


```{r cleancatch}

cleancatch_data <- catch_data %>%  
  mutate(site = str_to_sentence(site), # Standardizing site names
         site = str_trim(site), 
         site = str_replace_all(site, pattern = " ", replacement = "_")) %>%  
  mutate(species = str_to_sentence(species), # Standardizing species
         species = str_trim(species), 
         species = str_replace_all(species, pattern = " ", replacement = "_")) %>% 
  rename(common_name = species) %>% #to merge with species dictionary
  mutate(date = ymd(date)) %>% # Standardizing dates
  rename(timestamp = date) # Rename to merge with sonde data
  
print(cleancatch_data)

```


```{r cleansonde}
# Standardizing site names
sensor_data <- raw_sonde_data %>%  
  mutate(site = str_to_sentence(site), 
         site = str_trim(site), 
         site = str_replace_all(site, pattern = " ", replacement = "_")) %>% 
  mutate(timestamp = dmy_hm(timestamp)) %>% 
  mutate(turbidity = na_if(turbidity, -999))

head(sensor_data)

# Standardizing dates



```

```{r cleandictionary}
# Standardizing species
species_dictionary <- species_dictionary %>% 
  mutate(common_name = str_to_sentence(common_name), 
         common_name = str_trim(common_name), 
         common_name = str_replace_all(common_name, pattern = " ", replacement = "_"))
print(species_dictionary)
```
```{r merge}




```


2. A statistical summary table displaying the mean and standard deviation of fish counts and salinity for each species per zone.

3. A multi-panel plot (faceted by species) showing fish abundance along the salinity gradient, formatted to professional publication standards.

**Rules of Engagement**
Because this is an open-ended task, there is no single "correct" way to approach this and structure your code. However, think about all the data-wrangling and visualisation skills you have developed over recent weeks, and adhere to the following professional standards:

Monitor your AI agents: You are encouraged to use AI to help you write code. However, for every major data wrangling step, make sure you understand what is going on and are not just running AI-written code blindly.

Annotate your work: To help you with the above, make sure you include comments (#) explaining what the code is doing. If you used an AI prompt to generate a complex block of code, paste your prompt as a comment above it.

Utilise the tidyverse: There are many ways of wrangling data, but here we are looking for you to use your growing skills in dplyr, tidyr, stringr, and lubridate functions linked by the pipe (|> or %>%).

Reproducible workflow: Make sure you stay on top of your PC → Git → GitHub lifecycle, by utilising: save, commit, pull, push

**Mission Objectives**
*Phase 1:* Ingestion and Decontamination Load the four datasets. Before you attempt to join them, think about how you will need to wrangle and clean them individually:

Excel Tab Iteration: The catch log is split across multiple tabs (one for each site). You will need to figure out how to read all tabs and bind them together into a single data frame.

String Standardisation: The site and species names in the catch log are a mess (e.g., "MANGROVE JACK" vs "mangrove_jack"). Standardise all site and species columns to lowercase with underscores so they can securely match your dictionary and metadata tables later.

Temporal Parsing: The dates in the sonde dataset are messy character strings. Parse them into formal Date-Time objects.

Hardware Errors: The turbidity sensor occasionally failed and logged a value of -999.0. Identify these and convert them to true NA values before they ruin your averages.


*Phase 2:* The Relational Architecture Bring the data together into a cohesive format:

Summarise the water sensor data: The sensor data is logged every 15 minutes, but your catch data is daily. Create a daily summary table that calculates the mean temperature and salinity for each site on each day. (Hint: look at the floor_date() function to group by day).

Taxonomic Translation: Scientific journals will not accept "bream" or "flathead". Use a mutating join to merge your species_dictionary into your catch data, replacing the common names with accurate scientific taxonomy.

The Grand Join: Merge your translated catch data, your spatial metadata, and your daily water quality summaries into a single master data frame.


*Phase 3:* The Zero-Catch Framework The former student only recorded a row when they actually caught a fish:

Use tidyr::complete() to build a zero-data framework. Ensure that every species is represented at every site, on every sampling day.

Replace the resulting NA values in the catch column with 0 using coalesce().


*Phase 5:* Visual Communication Prove that your data rescue was successful. Construct a multi-panel plot using ggplot2 and facet_wrap().

The Goal: Visualise how the abundance of different species changes along the salinity gradient.

The Requirement: The plot should have professional axis labels, a clean theme (e.g., theme_minimal()), be ordered spatially (e.g. from upstream to downstream), and the facet labels (the scientific names) should really be italicised!

